---
title: 'MATH 324 Final Project'
author: "Amy Cao, Amy Folkestad, Anker Hojgaard and Jaxson Stathis"
output:
  pdf_document: default
  html_notebook: default
---

  
```{r, include = F, warning=F, message=F}
# front-matter
rm(list = ls()) #clear the workspace

library(Stat2Data)
library(tidyverse)
library(mosaic)
library(ggformula)
library(emmeans)
library(knitr)
library(kableExtra)
library(dplyr)


knitr::opts_chunk$set(echo = F)

# URL to the raw CSV file on GitHub
# this clean version has a series of dates that with odd formatting and missing year and month values as a result
# the dates were reformatted for the affected rows, months and years replaced the missing values

# Use the raw content URL instead of the repository page URL
file_url <- "https://raw.githubusercontent.com/BotanicalAmy/Stats-Project/main/hotel_booking_clean.csv"
hotel_bookings <- read.csv(file_url, header = TRUE)

```

```{r}
# new column GroupSize by adding Adults and Children
hotel_bookings$GroupSize <- hotel_bookings$Adults + hotel_bookings$Children

# new column StayLength by adding weekend and weeknights
hotel_bookings$StayLength <- hotel_bookings$Weekends + hotel_bookings$Weeknights

# Add HasChildren column
hotel_bookings$HasChildren <- ifelse(hotel_bookings$Children > 0, 1, 0)

# Preview the data
# head(hotel_bookings)

```

# 1. Project Description

*Should be in paragraph form (not bullets).  Written for a 3rd party client that may not have a strong statistical background. As part of your description, consider the following:*

Gold Mine Resorts is looking to understand the predictors for guests who cancel their reservations.The industry average for cancellation rates is twenty percent; however Gold Mine Resorts experienced a 32.8% cancellation rate for reservations between 2015 and 2018. Cancelled reservations represent a total of $4.3 million dollars in lost revenue for this time frame.

- *Description of the dataset*

The dataset includes 36,285 rows of booking data, spanning the course of four years (2015-2018). Data collected includes the number of guests (adult and children), the count of weekend and weeknight stays, guest upgrades (parking, meal plans, room types and count of special requests), booking details (price and reservation method), the date of stay, and whether or not the guest cancelled the reservation.

The status column (cancelled verses not cancelled), is the response variable of interest in this analysis.

- *Goal of study*

## 1.1 Research Objectives

The goal of this analysis is to understand the best predictors for guest cancellations in order to develop company policy to decrease the cancellation rate. Lowering the percent cancellation rate to 20% represents a revenue increase of approximately $2 million dollars over four years. 

*What are the overarching research objectives that the study is targeting? (not just a restatement of the original research objectives)*

**Objective 1:**  

The first objective will be to explore and define the most important quantitative variable for predicting whether a customer will cancel their reservation.

**Objective 2:**

The second objective will be to explore and define the most important categorical variable for predicting whether a customer will cancel their reservation.

**Objective 3:**

The final goal of our statistical analysis will combine our understanding of quantitative and categorical variables in order to create a combined statistical model to predict the customers who are most likely to cancel their reservations.


## 1.2 Variables

*What is (are) possible explanatory and response variables?*

The response variable of interest is status, whether or not a client cancelled their booking. The possible explanatory variables include both categoric and quantitative options. The number of requests from a client could have been classified as quantitative, but we selected to treat this as a categoric variable do to the clustering within the data. We made a similar to transformation to the children column, transforming this into a binary column indicating whether or not there were children in the group.


*Create a table here - it should include variable names and types*


|            | Description of Variable  | Variable Type |
|:----------:|:------------------------:|:-------------:|
| Status     |   Cancellation status of |       C       |
|            |   booking                |               |
|:----------:|:------------------------:|:-------------:|
| Meal       |   Type of meal in the    |       C       |
|            |   booking                |               |
|:----------:|:------------------------:|:-------------:|
| Parking    | Whether or not a parking |       C       |
|            | space is in the booking  |               |
|:----------:|:------------------------:|:-------------:|
| RoomType   |  Type of room booked     |       C       |
|:----------:|:------------------------:|:-------------:|
| Market     | Market segment of booking|       C       |
|:----------:|:------------------------:|:-------------:|
| Month      | Month of booking date    |       C       |
|:----------:|:------------------------:|:-------------:|
| Requests   | Num. of special requests |       C       |
|            | made by the customer     |               |
|:----------:|:------------------------:|:-------------:|
|HasChildren | Indicates whether or not |       C       |
|            | the booking has children |               |
|:----------:|:------------------------:|:-------------:| 
| Adults     |  Number of adults in the |       Q       |
|            |  booking                 |               |
|:----------:|:------------------------:|:-------------:|
| Children   | Number of children in the|       Q       |
|            | booking                  |               |
|:----------:|:------------------------:|:-------------:|
| Weekends   | Number of weekend nights |       Q       |
|            | in the booking           |               |
|:----------:|:------------------------:|:-------------:|
| Weeknights | Number of weeknights in  |       Q       |
|            | the booking              |               |
|:----------:|:------------------------:|:-------------:|
| LeadTime   | Number of days between   |       Q       |
|            | booking date and arrival |               |
|:----------:|:------------------------:|:-------------:|
| AvgPrice   | Average room price on the|       Q       |
|            | week of the booking      |               |
|:----------:|:------------------------:|:-------------:|
| Year       | Year of the reservation  |       Q       |
|:----------:|:------------------------:|:-------------:|




#  2. Detailed Exploratory Data Analysis (EDA)

*Spend time investigating the data thoroughly and use this section to present figures that allow you to visualize the relationship of the response to different predictors.  Include descriptions of the figures that illustrate what you can learn from them.*

```{r}
# quantitative EDA

# Compare means of each quantitative variable between canceled and not canceled groups
group_comparison <- hotel_bookings %>%
  group_by(Status) %>%
  summarize(
    Mean_LeadTime = mean(LeadTime, na.rm = TRUE),
    Mean_AvgPrice = mean(AvgPrice, na.rm = TRUE),
    Mean_GroupSize = mean(GroupSize, na.rm = TRUE),
    Mean_StayLength = mean(StayLength, na.rm = TRUE)
  )

# Round numeric columns to 3 decimal places
group_comparison_formatted <- group_comparison %>%
  mutate(across(where(is.numeric), ~round(., 3)))

# Create pretty kable table
kable(group_comparison_formatted, caption = "Quantitative Explanatory Variables") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#E8E8E8")


hotel_long <- hotel_bookings %>%
  select(Status, LeadTime, AvgPrice, GroupSize, StayLength) %>%
  pivot_longer(cols = c(LeadTime, AvgPrice, GroupSize, StayLength),
               names_to = "Variable",
               values_to = "Value")

# faceted boxplots for quantitative variables
ggplot(hotel_long, aes(x = Status, y = Value, fill = Status)) +
  geom_boxplot() +
  facet_wrap(~ Variable, scales = "free_y") +
  scale_fill_manual(values = c("Canceled" = "#FF6B6B", "Not_Canceled" = "#4ECDC4")) +
  labs(title = "Comparison of quantitative variables by cancellation status") +
  theme_minimal() 

```


```{r fig.width=6, fig.height=5}
# categorical EDA
# Create a bar chart with white background and title
gf_bar(~RoomType, fill=~Status, data = hotel_bookings) %>%
  gf_labs(
    title = "Cancellation by Room Type",
    x = "Room Type",
    y = "Count"
  ) %>%
  gf_refine(
    scale_fill_manual(values = c("Canceled" = "#FF6B6B", "Not_Canceled" = "#4ECDC4"))
  ) %>%
  gf_theme(
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom"
    )
  )

```

```{r fig.width=7, fig.height=5}

gf_bar(~Meal, fill=~Status, data = hotel_bookings) %>%
  gf_labs(
    title = "Cancellation by Meal Type",
    x = "Meal Type",
    y = "Count"
  ) %>%
  gf_refine(
    scale_fill_manual(values = c("Canceled" = "#FF6B6B", "Not_Canceled" = "#4ECDC4"))
  ) %>%
  gf_theme(
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      legend.position = "bottom"
    )
  )

gf_bar(~Market, fill=~Status, data = hotel_bookings) %>%
  gf_labs(
    title = "Cancellation by Market Segment",
    x = "Market Segment",
    y = "Count"
  ) %>%
  gf_refine(
    scale_fill_manual(values = c("Canceled" = "#FF6B6B", "Not_Canceled" = "#4ECDC4"))
  ) %>%
  gf_theme(
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      legend.position = "bottom"
    )
  )

```
```{r fig.width=8, fig.height=5}
# abbreviated month names in chronological order
month_order <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Convert Month to factor with levels in chronological order
hotel_bookings$Month <- factor(hotel_bookings$Month, levels = month_order)

# Create the bar chart with chronologically ordered months
gf_bar(~Month, fill=~Status, data = hotel_bookings) %>%
  gf_labs(
    title = "Cancellations by Month",
    x = "Month",
    y = "Count"
  ) %>%
  gf_refine(
    scale_fill_manual(values = c("Canceled" = "#FF6B6B", "Not_Canceled" = "#4ECDC4"))
  ) %>%
  gf_theme(
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      legend.position = "bottom"
    )
  )

gf_bar(~Requests, fill=~Status, data = hotel_bookings) %>%
  gf_labs(
    title = "Cancellations by Request Count",
    x = "Number of Special Requests",
    y = "Count"
  ) %>%
  gf_refine(
    scale_fill_manual(values = c("Canceled" = "#FF6B6B", "Not_Canceled" = "#4ECDC4"))
  ) %>%
  gf_theme(
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      legend.position = "bottom"
    )
  )
```
```{r fig.width=6, fig.height=5}
gf_bar(~HasChildren, fill=~Status, data = hotel_bookings) %>%
  gf_labs(
    title = "Cancellation by HasChildren",
    x = "Has Children (0 = No, 1 = Yes)",
    y = "Count"
  ) %>%
  gf_refine(
    scale_fill_manual(values = c("Canceled" = "#FF6B6B", "Not_Canceled" = "#4ECDC4"))
  ) %>%
  gf_theme(
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      legend.position = "bottom"
    )
  )

gf_bar(~Parking, fill=~Status, data = hotel_bookings) %>%
  gf_labs(
    title = "Cancellation by Parking Option",
    x = "Parking",
    y = "Count"
  ) %>%
  gf_refine(
    scale_fill_manual(values = c("Canceled" = "#FF6B6B", "Not_Canceled" = "#4ECDC4"))
  ) %>%
  gf_theme(
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      legend.position = "bottom"
    )
  )
```

```{r}
# Create a function to generate proportion tables
create_prop_table <- function(data, group_var) {
  group_var <- enquo(group_var)
  
  data %>%
    group_by(!!group_var, Status) %>%
    summarise(count = n(), .groups = "drop") %>%
    group_by(!!group_var) %>%
    mutate(proportion = count / sum(count))
}

# Now use the function with each variable
room_type_table <- create_prop_table(hotel_bookings, RoomType)
meal_table <- create_prop_table(hotel_bookings, Meal)
market_table <- create_prop_table(hotel_bookings, Market)
parking_table <- create_prop_table(hotel_bookings, Parking)
month_table <- create_prop_table(hotel_bookings, Month)
requests_table <- create_prop_table(hotel_bookings, Requests)
children_table <- create_prop_table(hotel_bookings, HasChildren)

# Function to create a nicely formatted kable table with rounded proportions
print_pretty_kable <- function(table_data, title) {
  table_data <- table_data %>%
    mutate(proportion = round(proportion, 3))
  
  kable(table_data, caption = title) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE) %>%
    column_spec(1, bold = TRUE) %>%
    row_spec(0, bold = TRUE, background = "#E8E8E8")
}

# Print each table with nice formatting
print_pretty_kable(room_type_table, "Room Type")
print_pretty_kable(meal_table, "Meal Option")
print_pretty_kable(market_table, "Market")
print_pretty_kable(parking_table, "Parking")
print_pretty_kable(month_table, "Month")
print_pretty_kable(requests_table, "Special Requests")
print_pretty_kable(children_table, "Has Children")

```

# 3. Statistical Analysis 

*Describe the statistical analysis that you used to answer Research Objectives 1-3. Each analysis should be labeled for the appropriate research objective. This section should summarize all relevant analyses that lead to your final conclusions/decisions/recommendations. Be sure to include:*

- *Model Assumptions & how checked/verified*
- *Interpretation of relevant estimates/statistics/p-values IN CONTEXT*
- *Any data manipulation to create new variables* 

*You should not include your code. However, I would like to see the code for final models that you are using to answer the research objectives in the appendix.* 

## 3.1 Objective 1

Model for lead time:
Coefficients:
              Estimate Std. Error z value Pr(>|z|)    
(Intercept)  1.8046260  0.0194124   92.96   <2e-16 ***
LeadTime    -0.0117484  0.0001593  -73.73   <2e-16 ***


Logit function for Status using LeadTime:
$$\widehat{log(\frac{\pi}{1-\pi})} = 1.8046260 - 0.0117484(LeadTime)$$

$$odds(Status) = e^{1.8046260 - 0.0117484(LeadTime)} $$

Probability function for Status using LeadTime:

$$p(Status) = \widehat \pi = \frac{e^{1.8046260 - 0.0117484(LeadTime)}}{1-e^{1.8046260 - 0.0117484(LeadTime)}}$$

$$odds \space ratio = e^{\widehat \beta_1} = e^{-0.0117484} = 0.988324$$ 
For every 1 day increase of LeadTime, the predicted odds of not cancelling the hotel booking decreases by a factor of 0.988. To further Gold Mine Resorts understanding of these statistics, Visuals of the probabilities and odds of cancellation are included below.

$$H_0: \beta_1 = 0, H_a: \beta_1 \neq 0$$
With a p-value < 2e^-16, we have enough evidence to reject the null hypothesis in favor of the alternative. LeadTime is useful for predicting hotel cancellations.



```{r}

# quantitative predictors
price_model <- glm(factor(Status) ~ AvgPrice, family = binomial(link = "logit"), data = hotel_bookings)
summary(price_model)

leadtime_model <- glm(factor(Status) ~ LeadTime, family = binomial(link = "logit"), data = hotel_bookings)
summary(leadtime_model)

staylength_model <- glm(factor(Status) ~ StayLength, family = binomial(link = "logit"), data = hotel_bookings)
summary(staylength_model)

groupsize_model <- glm(factor(Status) ~ GroupSize, family = binomial(link = "logit"), data = hotel_bookings)
summary(groupsize_model)

request_model <- glm(factor(Status) ~ Requests, family = binomial(link = "logit"), data = hotel_bookings)
summary(request_model)

#ASSESS
#converts status to binary where Canceled = 1 and Not_Canceled = 0, checking for linearity
hotel_bookings$Status_binary <- ifelse(hotel_bookings$Status == "Canceled", 1, 0)
emplogitplot1(Status_binary ~ LeadTime, data = hotel_bookings, ngroups = "all")
plot(leadtime_model, which = 1)

#USE
summary(leadtime_model)

```

```{r}

# Parameters from your logistic regression model
intercept <- 1.8046260
lead_time_coef <- -0.0117484

# Function to calculate probability of cancellation
cancellation_probability <- function(lead_time) {
  logit <- intercept + (lead_time_coef * lead_time)
  probability <- 1 / (1 + exp(-logit))
  return(probability)
}

# Function to calculate odds of cancellation
cancellation_odds <- function(lead_time) {
  prob <- cancellation_probability(lead_time)
  odds <- prob / (1 - prob)
  return(odds)
}

# Create a sequence of lead times from 0 to 365 days
lead_times <- seq(0, 365, by = 1)

# Calculate probabilities and odds for each lead time
probabilities <- sapply(lead_times, cancellation_probability)
odds <- sapply(lead_times, cancellation_odds)

# Create a data frame for plotting
plot_data <- data.frame(
  LeadTime = lead_times,
  Probability = probabilities,
  Odds = odds
)

# Create visualization
# Set up a 2x1 plotting layout
par(mfrow = c(2, 1), mar = c(4, 4, 2, 2))

# Plot 1: Probability vs Lead Time
plot(plot_data$LeadTime, plot_data$Probability, 
     type = "l", col = "royalblue", lwd = 2,
     xlab = "Lead Time (days)", ylab = "Probability of Cancellation",
     main = "Probability of Cancellation vs Lead Time", 
     ylim = c(0, 1))
grid()

# Add reference points
reference_points <- c(0, 30, 60, 90, 180, 365)
reference_probs <- sapply(reference_points, cancellation_probability)
points(reference_points, reference_probs, pch = 19, col = "#FF6B6B")

# Add labels for probability reference points
text(reference_points, reference_probs, 
     labels = paste0(round(reference_probs*100, 1), "%"), 
     pos = 3, col = "gray11")

# Plot 2: Odds vs Lead Time
plot(plot_data$LeadTime, plot_data$Odds, 
     type = "l", col = "midnightblue", lwd = 2,
     xlab = "Lead Time (days)", ylab = "Odds of Cancellation",
     main = "Odds of Cancellation vs Lead Time",
     ylim = c(0, 6))
grid()

# Add reference points for odds
reference_odds <- sapply(reference_points, cancellation_odds)
points(reference_points, reference_odds, pch = 19, col = "#FF6B6B")

# Add labels for odds reference points
text(reference_points, reference_odds, 
     labels = paste0(round(reference_odds, 2)), 
     pos = 3, col = "gray11")

# Reset plotting parameters
par(mfrow = c(1, 1))

# Additional information: Compare probability and odds changes
reference_table <- data.frame(
  LeadTime = reference_points,
  Probability = round(reference_probs, 4),
  ProbabilityPercentage = paste0(round(reference_probs*100, 1), "%"),
  Odds = round(reference_odds, 4)
)

# Print reference table
# print(reference_table)

# Calculate and display odds ratios for different intervals
base_odds <- cancellation_odds(0)
odds_ratios <- sapply(reference_points[-1], function(lt) {
  cancellation_odds(lt) / base_odds
})

cat("\nPercentage change in odds from Lead Time = 0:\n")
for (i in 1:length(odds_ratios)) {
  pct_change <- (odds_ratios[i] - 1) * 100
  cat(sprintf("Lead Time = %d days: %.1f%% change in odds\n", 
              reference_points[i+1], pct_change))
}

# Calculate and display odds ratio for one-day change
one_day_odds_ratio <- exp(lead_time_coef)
cat(sprintf("\nOdds ratio for one-day increase in lead time: %.4f\n", one_day_odds_ratio))
cat(sprintf("This corresponds to a %.2f%% decrease in odds per day\n", 
            (1 - one_day_odds_ratio) * 100))

```

## 3.2 Objective 2


```{r}

mod1 <- glm(factor(Status) ~ RoomType, family = binomial(link = "logit"), data = hotel_bookings)
summary(mod1)

plot(mod1, which = 1)

mod2 <- glm(factor(Status) ~ Parking, family = binomial(link = "logit"), data = hotel_bookings)
summary(mod2)

plot(mod2, which = 1)

mod3 <- glm(factor(Status) ~ Market, family = binomial(link = "logit"), data = hotel_bookings)
summary(mod3)

plot(mod3, which = 1)

mod4 <- glm(factor(Status) ~ Meal, family = binomial(link = "logit"), data = hotel_bookings)
summary(mod4)

plot(mod4, which = 1)

mod5 <- glm(factor(Status) ~ Month, family = binomial(link = "logit"), data = hotel_bookings)
summary(mod5)
confint.default(mod5)

plot(mod5, which = 1)

```

## 3.3 Objective 3

# 4. Conclusions 

*Succinct response to each question laid out in 1.1.  This a much shorter version of section 3, and focuses on conclusions rather than the analyses.*

**Objective 1:**  

```{r}
#gets the confidence interval, which does not include 0 therefore valid
confint.default(leadtime_model)
#gets the CI on the odds ratio scale, for every 1 unit increase in LeadTime, the odds of cancellation
#(vs. non-cancellation) multiply by 0.9880 to 0.9881 w/95% confidence
exp(confint.default(leadtime_model))

```

**Objective 2:**

```{r}


```


**Objective 3:**

*(all research objectives should be addressed, in the same order)*

Add in note on years, 2018 had a 36.7% cancellation rate, 2017 had a 14.8% cancellation rate
4.57 times as many bookings in 2018

# 5. Appendix 

*This section should include the code and output for the models used to answer the research objectives.*


```{r}
# add table
price_model <- glm(factor(Status) ~ AvgPrice, family = binomial(link = "logit"), data = hotel_bookings)
summary(price_model)

leadtime_model <- glm(factor(Status) ~ LeadTime, family = binomial(link = "logit"), data = hotel_bookings)
summary(leadtime_model)

staylength_model <- glm(factor(Status) ~ StayLength, family = binomial(link = "logit"), data = hotel_bookings)
summary(staylength_model)

groupsize_model <- glm(factor(Status) ~ GroupSize, family = binomial(link = "logit"), data = hotel_bookings)
summary(groupsize_model)
```