---
title: 'MATH 324 Final Project'
author: "Amy Cao, Amy Folkestad, Anker Hojgaard and Jaxson Stathis"
output:
  pdf_document: default
  html_notebook: default
---

  
```{r, include = F, warning=F, message=F}
# front-matter
rm(list = ls()) #clear the workspace

library(Stat2Data)
library(tidyverse)
library(mosaic)
library(ggformula)
library(emmeans)


knitr::opts_chunk$set(echo = F)

# URL to the raw CSV file on GitHub
# this clean version has a series of dates that with odd formatting and missing year and month values as a result
# the dates were reformatted for the affected rows, months and years replaced the missing values

# Use the raw content URL instead of the repository page URL
file_url <- "https://raw.githubusercontent.com/BotanicalAmy/Stats-Project/main/hotel_booking_cleaned.csv"
hotel_bookings <- read.csv(file_url, header = TRUE)

```

```{r}
# new column GroupSize by adding Adults and Children
hotel_bookings$GroupSize <- hotel_bookings$Adults + hotel_bookings$Children

# new column StayLength by adding weekend and weeknights
hotel_bookings$StayLength <- hotel_bookings$Weekends + hotel_bookings$Weeknights

# Add HasChildren column
hotel_bookings$HasChildren <- ifelse(hotel_bookings$Children > 0, 1, 0)

# Preview the data
 head(hotel_bookings)

```

# 1. Project Description

*Should be in paragraph form (not bullets).  Written for a 3rd party client that may not have a strong statistical background. As part of your description, consider the following:*

Gold Mine Resorts is looking to understand the predictors for guests who cancel their reservations.The industry average for cancellation rates is twenty percent; however Gold Mine Resorts experienced a 32.8% cancellation rate for reservations between 2015 and 2018. Cancelled reservations represent a total of $4.3 million dollars in lost revenue for this time frame.

- *Description of the dataset*

The dataset includes 36,285 rows of booking data, spanning the course of four years (2015-2018). Data collected includes the number of guests (adult and children), the count of weekend and weeknight stays, guest upgrades (parking, meal plans, room types and count of special requests), booking details (price and reservation method), the date of stay, and whether or not the guest cancelled the reservation.

The status column (cancelled verses not cancelled), is the response variable of interest in this analysis.

- *Goal of study*

## 1.1 Research Objectives

The goal of this analysis is to understand the best predictors for guest cancellations in order to develop company policy to decrease the cancellation rate. Lowering the percent cancellation rate to 20% represents a revenue increase of approximately $2 million dollars over four years. 

*What are the overarching research objectives that the study is targeting? (not just a restatement of the original research objectives)*

**Objective 1:**  

**Objective 2:**

**Objective 3:**


## 1.2 Variables

*What is (are) possible explanatory and response variables?*

The response variable of interest is status, whether or not a client cancelled their booking. The possible explanatory variables include both categoric and quantitative options. The number of requests from a client could have been classified as quantitative, but we selected to treat this as a categoric variable do to the clustering within the data. We made a similar to transformation to the children column, transforming this into a binary column indicating whether or not there were children in the group.


*Create a table here - it should include variable names and types*


|            | Description of Variable  | Variable Type |
|:----------:|:------------------------:|:-------------:|
| Status     |   Cancellation status of |       C       |
|            |   booking                |               |
|:----------:|:------------------------:|:-------------:|
| Meal       |   Type of meal in the    |       C       |
|            |   booking                |               |
|:----------:|:------------------------:|:-------------:|
| Parking    | Whether or not a parking |       C       |
|            | space is in the booking  |               |
|:----------:|:------------------------:|:-------------:|
| RoomType   |  Type of room booked     |       C       |
|:----------:|:------------------------:|:-------------:|
| Market     | Market segment of booking|       C       |
|:----------:|:------------------------:|:-------------:|
| Month      | Month of booking date    |       C       |
|:----------:|:------------------------:|:-------------:|
| Requests   | Num. of special requests |       C       |
|            | made by the customer     |               |
|:----------:|:------------------------:|:-------------:|
|HasChildren | Indicates whether or not |       C       |
|            | the booking has children |               |
|:----------:|:------------------------:|:-------------:| 
| Adults     |  Number of adults in the |       Q       |
|            |  booking                 |               |
|:----------:|:------------------------:|:-------------:|
| Children   | Number of children in the|       Q       |
|            | booking                  |               |
|:----------:|:------------------------:|:-------------:|
| Weekends   | Number of weekend nights |       Q       |
|            | in the booking           |               |
|:----------:|:------------------------:|:-------------:|
| Weeknights | Number of weeknights in  |       Q       |
|            | the booking              |               |
|:----------:|:------------------------:|:-------------:|
| LeadTime   | Number of days between   |       Q       |
|            | booking date and arrival |               |
|:----------:|:------------------------:|:-------------:|
| AvgPrice   | Average room price on the|       Q       |
|            | week of the booking      |               |
|:----------:|:------------------------:|:-------------:|
| Year       | Year of the reservation  |       Q       |
|:----------:|:------------------------:|:-------------:|




#  2. Detailed Exploratory Data Analysis (EDA)

*Spend time investigating the data thoroughly and use this section to present figures that allow you to visualize the relationship of the response to different predictors.  Include descriptions of the figures that illustrate what you can learn from them.*

```{r}
# quantitative EDA

# Compare means of each quantitative variable between canceled and not canceled groups
group_comparison <- hotel_bookings %>%
  group_by(Status) %>%
  summarize(
    Mean_LeadTime = mean(LeadTime, na.rm = TRUE),
    Mean_AvgPrice = mean(AvgPrice, na.rm = TRUE),
    Mean_GroupSize = mean(GroupSize, na.rm = TRUE),
    Mean_StayLength = mean(StayLength, na.rm = TRUE)
  )

print(group_comparison)

hotel_long <- hotel_bookings %>%
  select(Status, LeadTime, AvgPrice, GroupSize, StayLength) %>%
  pivot_longer(cols = c(LeadTime, AvgPrice, GroupSize, StayLength),
               names_to = "Variable",
               values_to = "Value")

# faceted boxplots for quantitative variables
ggplot(hotel_long, aes(x = Status, y = Value, fill = Status)) +
  geom_boxplot() +
  facet_wrap(~ Variable, scales = "free_y") +
  scale_fill_manual(values = c("Canceled" = "#FF6B6B", "Not_Canceled" = "#4ECDC4")) +
  labs(title = "Comparison of quantitative variables by cancellation status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
# categorical EDA
gf_bar(  ~RoomType ,  fill=~Status, data = hotel_bookings )
gf_bar(  ~Meal ,  fill=~Status, data = hotel_bookings )
gf_bar(  ~Market ,  fill=~Status, data = hotel_bookings )
gf_bar(  ~Parking ,  fill=~Status, data = hotel_bookings )
gf_bar(~Month, fill=~Status, data = hotel_bookings)
gf_bar(  ~Requests ,  fill=~Status, data = hotel_bookings )
gf_bar(  ~HasChildren, fill=~Status, data = hotel_bookings)


hotel_bookings %>%
  group_by(RoomType, Status) %>%
  summarise(count = n()) %>%
  group_by(RoomType) %>%
  mutate(proportion = count / sum(count))

hotel_bookings %>%
  group_by(Meal, Status) %>%
  summarise(count = n()) %>%
  group_by(Meal) %>%
  mutate(proportion = count / sum(count))

hotel_bookings %>%
  group_by(Market, Status) %>%
  summarise(count = n()) %>%
  group_by(Market) %>%
  mutate(proportion = count / sum(count))

hotel_bookings %>%
  group_by(Parking, Status) %>%
  summarise(count = n()) %>%
  group_by(Parking) %>%
  mutate(proportion = count / sum(count))

hotel_bookings %>%
  group_by(Month, Status) %>%
  summarise(count = n()) %>%
  group_by(Month) %>%
  mutate(proportion = count / sum(count))


hotel_bookings %>%
  group_by(Requests, Status) %>%
  summarise(count = n()) %>%
  group_by(Requests) %>%
  mutate(proportion = count / sum(count))

```

# 3. Statistical Analysis 

*Describe the statistical analysis that you used to answer Research Objectives 1-3. Each analysis should be labeled for the appropriate research objective. This section should summarize all relevant analyses that lead to your final conclusions/decisions/recommendations. Be sure to include:*

- *Model Assumptions & how checked/verified*
- *Interpretation of relevant estimates/statistics/p-values IN CONTEXT*
- *Any data manipulation to create new variables* 

*You should not include your code. However, I would like to see the code for final models that you are using to answer the research objectives in the appendix.* 

## 3.1 Objective 1

```{r}

# quantitative predictors
price_model <- glm(factor(Status) ~ AvgPrice, family = binomial(link = "logit"), data = hotel_bookings)
summary(price_model)

leadtime_model <- glm(factor(Status) ~ LeadTime, family = binomial(link = "logit"), data = hotel_bookings)
summary(leadtime_model)

staylength_model <- glm(factor(Status) ~ StayLength, family = binomial(link = "logit"), data = hotel_bookings)
summary(staylength_model)

groupsize_model <- glm(factor(Status) ~ GroupSize, family = binomial(link = "logit"), data = hotel_bookings)
summary(groupsize_model)

request_model <- glm(factor(Status) ~ Requests, family = binomial(link = "logit"), data = hotel_bookings)
summary(request_model)

#ASSESS
#converts status to binary where Canceled = 1 and Not_Canceled = 0, checking for linearity
hotel_bookings$Status_binary <- ifelse(hotel_bookings$Status == "Canceled", 1, 0)
emplogitplot1(Status_binary ~ LeadTime, data = hotel_bookings, ngroups = "all")
plot(leadtime_model, which = 1)

#USE
summary(leadtime_model)

```


## 3.2 Objective 2


```{r}

mod1 <- glm(factor(Status) ~ RoomType, family = binomial(link = "logit"), data = hotel_bookings)
summary(mod1)

plot(mod1, which = 1)

mod2 <- glm(factor(Status) ~ Parking, family = binomial(link = "logit"), data = hotel_bookings)
summary(mod2)

plot(mod2, which = 1)

mod3 <- glm(factor(Status) ~ Market, family = binomial(link = "logit"), data = hotel_bookings)
summary(mod3)

plot(mod3, which = 1)

mod4 <- glm(factor(Status) ~ Meal, family = binomial(link = "logit"), data = hotel_bookings)
summary(mod4)

plot(mod4, which = 1)

mod5 <- glm(factor(Status) ~ Month, family = binomial(link = "logit"), data = hotel_bookings)
summary(mod5)
confint.default(mod5)

plot(mod5, which = 1)

```

## 3.3 Objective 3

# 4. Conclusions 

*Succinct response to each question laid out in 1.1.  This a much shorter version of section 3, and focuses on conclusions rather than the analyses.*

**Objective 1:**  

```{r}
#gets the confidence interval, which does not include 0 therefore valid
confint.default(leadtime_model)
#gets the CI on the odds ratio scale, for every 1 unit increase in LeadTime, the odds of cancellation
#(vs. non-cancellation) multiply by 0.9880 to 0.9881 w/95% confidence
exp(confint.default(leadtime_model))

```

**Objective 2:**

```{r}


```


**Objective 3:**

*(all research objectives should be addressed, in the same order)*

# 5. Appendix 

*This section should include the code and output for the models used to answer the research objectives.*


```{r}
# quantitative predictors


```

```{r}
# categorical predictors

```

*Create a table here - it should include variable names and types*

```{r}
# add table
price_model <- glm(factor(Status) ~ AvgPrice, family = binomial(link = "logit"), data = hotel_bookings)
summary(price_model)

leadtime_model <- glm(factor(Status) ~ LeadTime, family = binomial(link = "logit"), data = hotel_bookings)
summary(leadtime_model)

staylength_model <- glm(factor(Status) ~ StayLength, family = binomial(link = "logit"), data = hotel_bookings)
summary(staylength_model)

groupsize_model <- glm(factor(Status) ~ GroupSize, family = binomial(link = "logit"), data = hotel_bookings)
summary(groupsize_model)
```


#  2. Detailed Exploratory Data Analysis (EDA)

*Spend time investigating the data thoroughly and use this section to present figures that allow you to visualize the relationship of the response to different predictors.  Include descriptions of the figures that illustrate what you can learn from them.*

```{r}
# quantitative EDA

#ASSESS
#converts status to binary where Canceled = 1 and Not_Canceled = 0, checking for linearity
hotel_bookings$Status_binary <- ifelse(hotel_bookings$Status == "Canceled", 1, 0)
emplogitplot1(Status_binary ~ LeadTime, data = hotel_bookings, ngroups = "all")
plot(leadtime_model, which = 1)

#USE
summary(leadtime_model)

#gets the confidence interval, which does not include 0 therefore valid
confint.default(leadtime_model)
#gets the CI on the odds ratio scale, for every 1 unit increase in LeadTime, the odds of cancellation
#(vs. non-cancellation) multiply by 0.9880 to 0.9881 w/95% confidence
exp(confint.default(leadtime_model))

```

```{r}
# categorical EDA

```

# 3. Statistical Analysis 

*Describe the statistical analysis that you used to answer Research Objectives 1-3. Each analysis should be labeled for the appropriate research objective. This section should summarize all relevant analyses that lead to your final conclusions/decisions/recommendations. Be sure to include:*

- *Model Assumptions & how checked/verified*
- *Interpretation of relevant estimates/statistics/p-values IN CONTEXT*
- *Any data manipulation to create new variables* 

*You should not include your code. However, I would like to see the code for final models that you are using to answer the research objectives in the appendix.* 

## 3.1 Objective 1

## 3.2 Objective 2

## 3.3 Objective 3

# 4. Conclusions 

*Succinct response to each question laid out in 1.1.  This a much shorter version of section 3, and focuses on conclusions rather than the analyses.*

**Objective 1:**  

**Objective 2:**

**Objective 3:**

*(all research objectives should be addressed, in the same order)*

# 5. Appendix 

*This section should include the code and output for the models used to answer the research objectives.*
